---
title: "Life Expectancy"
author: "Dillan Sant, Chase Mathis"
format: html
editor: visual
execute: 
  echo: false
  message: false
  warning: false
---

```{r}
#| label: packages
#| echo: false
#| message: false


library(tidyverse)
library(tidymodels)
library(knitr)
library(patchwork)
library(broom)
library(MASS)
library(fuzzyjoin)
#install.packages("fuzzyjoin")

life <- read_csv("data/life_exp.csv")
ggplot2::theme_set(ggplot2::theme_bw())
worldData <- map_data("world")

```

```{r}
## fixing for graph plots
worldData <- worldData |> 
  mutate(region = case_when(
    region == "Antigua" ~ "Antigua and Barbuda",
    region == "Bolivia" ~ "Bolivia (Plurinational State of)",
    region == "Brunei" ~ "Brunei Darussalam",
    region == "Ivory Coast" ~ "Côte d'Ivoire",
    region == "Cape Verde" ~ "Cabo Verde",
    region == "Democratic Republic of the Congo" ~ "Congo",
    region == "Czech Republic" ~ "Czechia",
    region == "North Korea" ~ "Democratic People's Republic of Korea",
    region == "Iran" ~ "Iran (Islamic Republic of)",
    region == "Laos" ~ "Lao People's Democratic Republic",
    region == "Micronesia" ~ "Micronesia (Federated States of)",
    region == "South Korea" ~ "Republic of Korea",
    region == "Moldova" ~ "Republic of Moldova",
    region == "Russia" ~ "Russian Federation",
    region == "Syria" ~ "Syrian Arab Republic",
    region == "Trinidad" ~ "Trinidad and Tobago",
    region == "UK" ~ "United Kingdom of Great Britain and Northern Ireland",
    region == "Tanzania" ~ "United Republic of Tanzania",
    region == "USA" ~ "United States of America",
    region == "Venezuela" ~ "Venezuela (Bolivarian Republic of)",
    region == "Vietnam" ~ "Viet Nam",
    region == "Saint Kitts" ~ "Saint Kitts and Nevis",                
    region != "" ~ region
  ))
```

```{r}
life <- life %>% mutate(Status = factor(Status, levels = c("Developing", "Developed"))) %>%
  rename(life_exp = "Life expectancy", adult_mortality = "Adult Mortality", 
        infant_mortality = "infant deaths", pct_expend = "percentage expenditure", 
        hep_b = "Hepatitis B", under_five_deaths = "under-five deaths", 
        tot_expend = "Total expenditure", hiv_aids = "HIV/AIDS",
        thinness_adole = "thinness  1-19 years", thinness_infant = "thinness 5-9 years", 
        income_comp = "Income composition of resources") 
```

```{r}
#| label: pre-processing
set.seed(123)
split <- initial_split(life)
training_life <- training(split)
testing_life <- testing(split)

```

```{r}
life |> 
  filter(Year == 2015) |> 
  dplyr::select(Country,life_exp) |> 
  arrange(desc(life_exp)) |> 
  slice(1:10) |> 
  kable(col.names = c("Country", "Life Expectancy"), caption = "Life Expectancy of the Top 10 Countries")
```

## Overview of Data

### Controllable Variables

-   `Alcohol`: Alcohol consumption per capita (liters of pure alcohol)
-   `pct_expend`: Expenditure on health as a percentage of GDP per capita
-   `hep_b`: Hepatitis B immunization rate among 1-year-olds (%)
-   `Polio`: Polio immunization rate among 1-year-olds (%)
-   `tot_expend`: Government expenditure on healthcare as a percentage of total government expenditure
-   `Diphtheria`: Diphtheria tetnus toxoid and pertussis immunization rate among 1-year-olds (%)
-   `hiv_aids`: Number of deaths of 0-4 year-olds from HIV/AIDS per 1000 live births
-   `Schooling`: Average number of years of schooling

### Uncontrollable Variables (Nuiscance Variables)

-   `BMI`: Average BMI (Body Mass Index) of entire country's population
-   `GDP`: GDP per capita
-   `Population`: Total population of country
-   `thinnes_adole`: Prevalence of "thinness" among adolescents aged 10-19 (%)
-   `thinness_infant`: Prevalence of "thinness" among infants aged 5-9 (%)
-   `income_comp`: Human Development Index in terms of income composition of resources (0 to 1)
-   `Status`: Developmental status of country (Developed or Developing)

### Indicator Variables

-   `under_five_deaths`: Number of deaths of 5-year-olds or younger per 1000 people
-   `life_exp`: Average expectancy in country (years)
-   `adult_mortality`: Number of deaths of people aged 15-60 per 1000 people
-   `infant_mortality`: Number of infants deaths per 1000 infants

```{r}
control_vars <- c("Alcohol","pct_expend", "hep_b", "Polio", "tot_expend", "Diphtheria", "hiv_aids", "Schooling")
uncontrol_vars <- c("BMI", "GDP", "Population", "thinnes_adole", "thinness_infant", "income_comp", "Status")
indicator_vars <- c("under_five_deaths", "life_exp", "adult_mortality", "infant_mortality")
```

## Exploratory Data Analysis

```{r}

#recipe <- recipe(life) |> 
#  update_role(Country, new_role = "id variable") |> 
#  step_mutate(Country = as.character(Country)) |> 
#  step_dummy(all_nominal()) |> 
#  prep()
#
#life <- bake(recipe, life)
#
#
#
#p2 <- ggplot(life, aes(x = life_exp)) + 
#  geom_histogram(bins = 20)+
#  labs(
#    x = "Life Expectancy",
#    y = "County",
#    title = "Distribution of Life Expectancy",
#    caption = "Notice that the distribution is Left Skewed"
#  )


```

```{r}
life %>% filter(Year == 2015) %>% ggplot(aes(x = GDP, y = life_exp, color = Status)) + 
  geom_point()
```


```{r}
deltaLifeExp <- life |> 
  filter(Year == 2000 | Year == 2015) |> 
  group_by(Country) |> 
  mutate(delta_Life_Exp = lag(life_exp) - life_exp) |> 
  relocate(delta_Life_Exp) |> 
  ungroup() |> 
  drop_na(delta_Life_Exp) 

deltaLifeExp |> 
  ggplot(aes(x = delta_Life_Exp)) + 
  geom_histogram(bins = 10) + 
  labs(
    x = "Change in Life Expectancy",
    y = "Frequency"
  )

life |> 
  ggplot(aes(x = Year, y = life_exp, group = Year)) + 
  geom_boxplot()

```

```{r}

worldData |> 
full_join(deltaLifeExp, by = c("region" = "Country")) |> 
ggplot(aes(x = long, y = lat, group = group, fill = delta_Life_Exp)) +
  geom_polygon() + 
  scale_fill_continuous(type = "viridis")
  

```

Generally countries increased life expectancy, especially in Sub-Saharan Africa. Which countries experienced a decrease in life expectancy.

```{r}
deltaLifeExp %>% filter(delta_Life_Exp < 0) %>% 
  dplyr::select(delta_Life_Exp, Country) %>% arrange(delta_Life_Exp) %>%
  rename("Change in Life Expectancy" = delta_Life_Exp) %>% kable()
```

From 2000-2015, the nations that experienced a decrease in life expectancy are Syria, St. Vincent and the Grenandines, Libya, Paraguay, Yemen, Romania, Iraq, Estonia, and Grenada. All of these nations except for Romania are developing.

```{r}
negativeLife <- deltaLifeExp |> 
  filter(delta_Life_Exp < 0)
```

## Research Question 1: Given a country is developing, what can they do to increase their Life Expectancy?

```{r}
#| label: data-pre-processing

life_control_developing <- recipe( ~ ., data = training_life) |> 
  update_role(Country, new_role = "ID") |> 
  step_filter(Status == "Developing") |>   # only developing countries
  step_select(all_of(control_vars), Country, life_exp, skip = TRUE) |>    #only uses control variables
    update_role(life_exp, new_role = "outcome")


life_uncontrol_developing <- recipe(life_exp ~ ., data = training_life) |> 
  update_role(Country, new_role = "ID") |> 
  step_filter(Status == "Developing") |>   # only developing countries
  step_select(all_of(uncontrol_vars), Country) |> #only uses control variables
  step_dummy(all_nominal_predictors())
```

```{r}
#| label: models

linear_spec <- linear_reg() |> 
  set_engine("lm")
```

```{r}
wflow_control_developed <- workflow() %>%
  add_model(linear_spec) %>%
  add_recipe(life_control_developing) 


control_developed_fit <- wflow_control_developed |> 
  fit(data = training_life)

tidy(control_developed_fit) |>
  mutate(significant = if_else(p.value < 0.05, "Significant", "Not Significant")) |> 
  dplyr::select(term, estimate, p.value, significant) |> 
  arrange(desc(abs(estimate))) |> 
  kable()
```

### Include in first research question: For the small subset of countries that saw a decrease in life expectancy from 2000-2015, what factors led to this decrease in life expectancy?

```{r}
delta_cont <- life |> 
  filter(Year == 2000 | Year == 2015) |> 
  group_by(Country) |> 
  mutate(delta_Life_Exp = lag(life_exp) - life_exp, delta_alc = lag(Alcohol) - Alcohol, delta_pe = lag(pct_expend) - pct_expend, delta_hep = lag(hep_b) - hep_b, delta_pol = lag(Polio) - Polio, delta_expend = lag(tot_expend) - tot_expend, delta_dip = lag(Diphtheria) - Diphtheria, delta_hiv = lag(hiv_aids) - hiv_aids, delta_school = lag(Schooling) - Schooling) |>
  ungroup()
```

## Research Question 2: For countries that already have a high life expectancy, is it economically beneficial to attempt to marginally increase life expectancy?

## Conclusion
- suggest in which factors specific nations should invest their money in based on modeling during both research questions, or, suggest not to increase investment in health care for nations with already high life expectancy,  

