---
title: "Life Expectancy"
author: "Dillan Sant, Chase Mathis"
format: html
editor: visual
execute: 
  echo: false
---

```{r}
#| label: packages
#| echo: false
#| message: false


library(tidyverse)
library(tidymodels)
library(knitr)
library(patchwork)
library(broom)
library(MASS)
library(fuzzyjoin)
#install.packages("fuzzyjoin")

life <- read_csv(file = here::here("data/life_exp.csv"))
ggplot2::theme_set(ggplot2::theme_bw())
worldData <- map_data("world")

```

```{r}
## annoying fixing
worldData <- worldData |> 
  mutate(region = case_when(
    region == "Antigua" ~ "Antigua and Barbuda",
    region == "Bolivia" ~ "Bolivia (Plurinational State of)",
    region == "Brunei" ~ "Brunei Darussalam",
    region == "Ivory Coast" ~ "Côte d'Ivoire",
    region == "Cape Verde" ~ "Cabo Verde",
    region == "Democratic Republic of the Congo" ~ "Congo",
    region == "Czech Republic" ~ "Czechia",
    region == "North Korea" ~ "Democratic People's Republic of Korea",
    region == "Iran" ~ "Iran (Islamic Republic of)",
    region == "Laos" ~ "Lao People's Democratic Republic",
    region == "Micronesia" ~ "Micronesia (Federated States of)",
    region == "South Korea" ~ "Republic of Korea",
    region == "Moldova" ~ "Republic of Moldova",
    region == "Russia" ~ "Russian Federation",
    region == "Syria" ~ "Syrian Arab Republic",
    region == "Trinidad" ~ "Trinidad and Tobago",
    region == "UK" ~ "United Kingdom of Great Britain and Northern Ireland",
    region == "Tanzania" ~ "United Republic of Tanzania",
    region == "USA" ~ "United States of America",
    region == "Venezuela" ~ "Venezuela (Bolivarian Republic of)",
    region == "Vietnam" ~ "Viet Nam",
    region == "Saint Kitts" ~ "Saint Kitts and Nevis",                
    region != "" ~ region
  ))
```

```{r}
life <- life %>% mutate(Status = factor(Status, levels = c("Developing", "Developed"))) %>%
  rename(life_exp = "Life expectancy", adult_mortality = "Adult Mortality", 
        infant_mortality = "infant deaths", pct_expend = "percentage expenditure", 
        hep_b = "Hepatitis B", under_five_deaths = "under-five deaths", 
        tot_expend = "Total expenditure", hiv_aids = "HIV/AIDS",
        thinness_adole = "thinness  1-19 years", thinness_infant = "thinness 5-9 years", 
        income_comp = "Income composition of resources") 
```

```{r}
life |> 
  filter(Year == 2015) |> 
  dplyr::select(Country,life_exp) |> 
  arrange(desc(life_exp)) |> 
  slice(1:10) |> 
  kable(col.names = c("Country", "Life Expectancy"), caption = "Life Expectancy of the Top 10 Countries")
```

## Overview of Data

### Controllable Variables

-   `Alcohol`: Alcohol consumption per capita (liters of pure alcohol)
-   `pct_expend`: Expenditure on health as a percentage of GDP per capita
-   `hep_b`: Hepatitis B immunization rate among 1-year-olds (%)
-   `Polio`: Polio immunization rate among 1-year-olds (%)
-   `tot_expend`: Government expenditure on healthcare as a percentage of total government expenditure
-   `Diphtheria`: Diphtheria tetnus toxoid and pertussis immunization rate among 1-year-olds (%)
-   `hiv_aids`: Number of deaths of 0-4 year-olds from HIV/AIDS per 1000 live births
-   `Schooling`: Average number of years of schooling

### Uncontrollable Variables (Nuiscance Variables)

-   `BMI`: Average BMI (Body Mass Index) of entire country's population
-   `GDP`: GDP per capita
-   `Population`: Total population of country
-   `thinnes_adole`: Prevalence of "thinness" among adolescents aged 10-19 (%)
-   `thinness_infant`: Prevalence of "thinness" among infants aged 5-9 (%)
-   `income_comp`: Human Development Index in terms of income composition of resources (0 to 1)
-   `Status`: Developmental status of country (Developed or Developing)

### Indicator Variables

-   `under_five_deaths`: Number of deaths of 5-year-olds or younger per 1000 people
-   `life_exp`: Average expectancy in country (years)
-   `adult_mortality`: Number of deaths of people aged 15-60 per 1000 people
-   `infant_mortality`: Number of infants deaths per 1000 infants

## Exploratory Data Analysis

```{r}

#recipe <- recipe(life) |> 
#  update_role(Country, new_role = "id variable") |> 
#  step_mutate(Country = as.character(Country)) |> 
#  step_dummy(all_nominal()) |> 
#  prep()
#
#life <- bake(recipe, life)
#
#
#
#p2 <- ggplot(life, aes(x = life_exp)) + 
#  geom_histogram(bins = 20)+
#  labs(
#    x = "Life Expectancy",
#    y = "County",
#    title = "Distribution of Life Expectancy",
#    caption = "Notice that the distribution is Left Skewed"
#  )


```

```{r}
deltaLifeExp <- life |> 
  filter(Year == 2000 | Year == 2015) |> 
  group_by(Country) |> 
  mutate(delta_Life_Exp = lag(life_exp) - life_exp) |> 
  relocate(delta_Life_Exp) |> 
  ungroup() |> 
  drop_na(delta_Life_Exp) 

deltaLifeExp |> 
  ggplot(aes(x = delta_Life_Exp)) + 
  geom_histogram(bins = 10) + 
  labs(
    x = "Change in Life Expectancy",
    y = "Frequency"
  )

life |> 
  ggplot(aes(x = Year, y = life_exp, group = Year)) + 
  geom_boxplot()

```

```{r}

worldData |> 
full_join(deltaLifeExp, by = c("region" = "Country")) |> 
ggplot(aes(x = long, y = lat, group = group, fill = delta_Life_Exp)) +
  geom_polygon() + 
  scale_fill_continuous(type = "viridis")
  

```

Generally countries increased life expectancy. Which ones did not?

```{r}
deltaLifeExp |> 
  filter(delta_Life_Exp < 0)
  
```
