---
title: "STA325 Final Project Proposal"
author: "Dillan Sant, Chase Mathis"
format: revealjs
execute: 
  warning: false
  echo: false
editor: visual
---



## Executive Summary

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
library(knitr)
library(patchwork)
ggplot2::theme_set(ggplot2::theme_bw())
life <- read_csv("data/life_exp.csv")
```

```{r}
life <- life %>% mutate(Status = factor(Status, levels = c("Developing", "Developed"))) %>%
  rename(life_exp = "Life expectancy", adult_mortality = "Adult Mortality", 
        infant_mortality = "infant deaths", pct_expend = "percentage expenditure", 
        hep_b = "Hepatitis B", under_five_deaths = "under-five deaths", 
        tot_expend = "Total expenditure", hiv_aids = "HIV/AIDS",
        thinness_adole = "thinness  1-19 years", thinness_infant = "thinness 5-9 years", 
        income_comp = "Income composition of resources")
```


1. Project objectives
2. High-level summary of project plans
3. Importance of project, who can it benefit
4. Other relevant info

## Data Description

- Data from the Global Healthy Observatory's (GHO) data repository, made public for global health data analysis
- Health factors collected from GHO repository, economic data collected from United Nations' website
- Each observation represents nation in a given year and includes health and economic data, as well as each nation's life expectancy for that year
- 22 variables:
  - 21 predictors (see next slides)
  - 1 response: life expectancy, in years

## Data Dictionary
- `Country`: Country name 
- `Year`: Year country's GHO data
- `Status`: Developmental status of country (Developed or Developing)
- `life_exp`: Average expectancy in country (years)
- `adult_mortality`: Number of deaths of people aged 15-60 per 1000 people
- `infant_mortality`: Number of infants deaths per 1000 infants

## Data Dictionary
- `Alcohol`: Alcohol consumption per capita (liters of pure alcohol)
- `pct_expend`: Expenditure on health as a percentage of GDP per capita
- `hep_b`: Hepatitis B immunization rate among 1-year-olds (%)
- `Measles`: Number of Measles cases reported per 1000 people
- `BMI`: Average BMI (Body Mass Index) of entire country's population 


## Data Dictionary
- `under_five_deaths`: Number of deaths of 5-year-olds or younger per 1000 people
- `Polio`: Polio immunization rate among 1-year-olds (%)
- `tot_expend`: Government expenditure on healthcare as a percentage of total government expenditure
- `Diphtheria`: Diphtheria tetnus toxoid and pertussis immunization rate among 1-year-olds (%)
- `hiv_aids`: Number of deaths of 0-4 year-olds from HIV/AIDS per 1000 live births

## Data Dictionary
- `GDP`: GDP per capita
- `Population`: Total population of country
- `thinnes_adole`: Prevalence of "thinness" among adolescents aged 10-19 (%)
- `thinness_infant`: Prevalence of "thinness" among infants aged 5-9 (%) 
- `income_comp`: Human Development Index in terms of income composition of resources (0 to 1)
- `Schooling`: Average number of years of schooling

## Roadmap

1. Inference + prediction goals
2. Possible modeling strategies
3. How we will select optimal model, best achieves our prediction/inference goals

Flowchart to outline key steps?

## Exploratory Data Analysis

plots, descriptive statistics, initial findings, etc.

