---
title: "STA325 Final Project Proposal"
author: "Dillan Sant, Chase Mathis"
format: revealjs
execute: 
  warning: false
  echo: false
editor: visual
---

## Executive Summary

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
library(tidymodels)
library(knitr)
library(patchwork)
library(Hmisc) # for histograms
library(DiagrammeR)
ggplot2::theme_set(ggplot2::theme_bw())
life <- read_csv("data/life_exp.csv")
```

```{r}
life <- life %>% mutate(Status = factor(Status, levels = c("Developing", "Developed"))) %>%
  rename(life_exp = "Life expectancy", adult_mortality = "Adult Mortality", 
        infant_mortality = "infant deaths", pct_expend = "percentage expenditure", 
        hep_b = "Hepatitis B", under_five_deaths = "under-five deaths", 
        tot_expend = "Total expenditure", hiv_aids = "HIV/AIDS",
        thinness_adole = "thinness  1-19 years", thinness_infant = "thinness 5-9 years", 
        income_comp = "Income composition of resources") 
```

-   Our objective is to discover which health or economic factors of a nation impacts the country's average life expectancy the most
-   We aim to craft a model that has high predictive power on life expectancy, but also allows for inferable results that could be interpreted by professionals in the field of global health

## Data Description

-   Data from the Global Healthy Observatory's (GHO) data repository, made public for global health data analysis
-   Health factors collected from GHO repository, economic data collected from United Nations' website
-   Each observation represents nation in a given year and includes health and economic data, as well as each nation's life expectancy for that year
-   22 variables:
    -   20 predictors (see next slides)
    -   1 response: life expectancy, in years

## Data Dictionary

-   `Year`: Year country's GHO data
-   `Status`: Developmental status of country (Developed or Developing)
-   `life_exp`: Average expectancy in country (years)
-   `adult_mortality`: Number of deaths of people aged 15-60 per 1000 people
-   `infant_mortality`: Number of infants deaths per 1000 infants

## Data Dictionary

-   `Alcohol`: Alcohol consumption per capita (liters of pure alcohol)
-   `pct_expend`: Expenditure on health as a percentage of GDP per capita
-   `hep_b`: Hepatitis B immunization rate among 1-year-olds (%)
-   `Measles`: Number of Measles cases reported per 1000 people
-   `BMI`: Average BMI (Body Mass Index) of entire country's population

## Data Dictionary

-   `under_five_deaths`: Number of deaths of 5-year-olds or younger per 1000 people
-   `Polio`: Polio immunization rate among 1-year-olds (%)
-   `tot_expend`: Government expenditure on healthcare as a percentage of total government expenditure
-   `Diphtheria`: Diphtheria tetnus toxoid and pertussis immunization rate among 1-year-olds (%)
-   `hiv_aids`: Number of deaths of 0-4 year-olds from HIV/AIDS per 1000 live births

## Data Dictionary

-   `GDP`: GDP per capita
-   `Population`: Total population of country
-   `thinnes_adole`: Prevalence of "thinness" among adolescents aged 10-19 (%)
-   `thinness_infant`: Prevalence of "thinness" among infants aged 5-9 (%)
-   `income_comp`: Human Development Index in terms of income composition of resources (0 to 1)
-   `Schooling`: Average number of years of schooling

## Roadmap

1.  Inference + prediction goals
2.  Possible modeling strategies
3.  How we will select optimal model, best achieves our prediction/inference goals

Flowchart to outline key steps?

## Inference and Prediction Goals

-   There are some important research questions that the dataset can help us answer. For instance:

    -   What impacts `Life Expectancy` the most?

    -   What is the relationship between `Life Expectancy` and the other predictors?

    -   Can we predict an increase or decrease in `Life Expectancy` in the future? i.e. Time Series Analysis?

    -   How well can we predict `Life Expectancy` given the data we have?

## Possible Modeling Strategies

1.  Lasso Regression:
    -   This model has many predictors. Can we select the most important ones using Lasso?
2.  Linear and Nonlinear Regression
3.  Boosting, Bagging, and general Trees:
    1.  Given that the relationship between Linear and Nonlinear Regression is not linear, can we use `Tree-based models` to predict?

## How Will we Select the Best Model?

-   We aim to select a model with good performance without sacrificing much inference
-   Our primary goals are to find what health or economic factors of a country impact their average life expectancy and allow for interpretable results that can be helpful to the field of global health
-   We will choose a model with relatively small test error and one whose results/predictions are interpretable
-   Thus, we will not necessarily choose the model with the lowest test MSE

## Roadmap

```{r}
DiagrammeR::grViz(diagram = "digraph flowchart {
  node [fontname = arial, shape = oval, color = Lavender, style = filled]
  tab1 [label = '@@1']
  tab2 [label = '@@2']
  tab3 [label = '@@3']
  tab4 [label = '@@4']
  tab5 [label = '@@5']
  tab6 [label = '@@6']
  
  tab1 -> tab2
  tab2 -> tab3
  tab3 -> tab4
  tab3 -> tab5
  tab4 -> tab6
  tab5 -> tab6
}
  
  [1]: 'Gather and clean data'
  [2]: 'Explore relationships between health and economic factors'    
  [3]: 'Model'  
  [4]: 'Choose model with high predictive power'
  [5]: 'Determine which factors most affect life expectancy'
  [6]: 'Use results to predict future life expectancy'
  ")
```


## Exploratory Data Analysis

```{r}

recipe <- recipe(life) |> 
  update_role(Country, new_role = "id variable") |> 
  step_mutate(Country = as.character(Country)) |> 
  step_dummy(all_nominal()) |> 
  prep()

life <- bake(recipe, life)

p1 <- corrr::correlate(life) |> 
  select(term, life_exp) |> 
  ggplot(aes(x = reorder(term,life_exp), y = life_exp, fill = life_exp)) + 
  geom_col() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  labs(
    x = "Term",
    y = "Correlation",
    fill = "Correlation",
    title = "Correlation with Life Expectancy"
  )

p2 <- ggplot(life, aes(x = life_exp)) + 
  geom_histogram(bins = 20) + 
  labs(
    x = "Life Expectancy",
    y = "County",
    title = "Distribution of Life Expectancy",
    caption = "Notice that the distribution is Left Skewed"
  )

corrr::correlate(life) |> 
  select(term, life_exp) |> 
  filter(term!= life_exp) |> 
  mutate(life_exp = abs(life_exp)) |> 
  arrange(desc(life_exp)) |> 
  head(2) |> 
  pull(term)
# what is relationship between the two must importnat factors
p3 <- ggplot(life, aes(x = Schooling, y = life_exp)) + 
  geom_point()
```

```{r}
#| warning: false
p1 + p2 + p3 + 
  plot_layout(ncol = 3)
```

```{r}
#| include: false
life |> 
  select(-Country,-Year) |> 
  gather() |> 
  ggplot(aes(x = value))+
  geom_density()+
  facet_wrap(~key, scales = "free")

```
